 

### 2023-05-08 markdown/methods.Rmd
missing v handling finished. batch correction started

#### Missing value handling
When analyzing multiple batches simultaneously, it is possible to encounter situations where a particular protein is not detected in one of the batches. From a biological perspective, it is highly unlikely that a cell completely lacks a single protein, as it would imply a complete loss of function for that protein. Therefore, an intensity value of 0 is close to impossible for most proteins. In order to handle this dark space in the expression set, several methods can be chosen in the application. 
A variety of function employs replacing the missing values with the mean or median of the rest of the matrix. This procedure introduces no bias in the data, although differential expression for the sample carrying the missing feature can not be expected as well. Even a more conservative approach is dropping the rows for the feature with the missing value in one of the observed samples. 
Another common practice is utilizing the K-nearest neighbor algorithm (=KNN), which imputes the missing values by a euclidean metrics of the neighboring values in the columns where the feature is not missing. $K$ denotes for the count of neighboring values of a gene and can be selected upon preference.

>>>>

#### Missing value handling
When analyzing multiple batches simultaneously, it is possible to encounter situations where a particular protein is not detected in one of the batches. From a biological perspective, it is highly unlikely that a cell completely lacks a single protein, as it would imply a complete loss of function for that protein. Therefore, an intensity value of 0 is close to impossible for most proteins. In order to handle this dark space in the expression set, several methods can be chosen in the application. 
A variety of function employs replacing the missing values with the mean or median of the rest of the matrix. This procedure introduces no bias in the data if the missingness is low, although differential expression for the sample carrying the missing feature can not be expected as well. Even a more conservative approach is dropping the rows for the feature with the missing value in one of the observed samples. 
A common practice in proteomics is utilizing the K-nearest neighbor algorithm (=KNN), which imputes the missing values by a euclidean metrics of the neighboring values in the columns where the feature is not missing. $K$ denotes for the count of neighboring values of a gene and can be selected upon preference. The euclidean distance between two points is defined as $d = \sqrt{a^2 * b^2}$, the $K$ nearest neighbors average value will be assigned to the intensity of the missing feature. Since the average is a statistical metric which can be biased easily by skewed distributions, normalization is an important task before performing missing value imputation. KNN was benchmarked  for the regression of protein abundance intensities and showed an Area under the Curve (AUC) above 0.7 when testing on a human proteome test set \citep{Lan2013}.

#### Batch correction 
Since mass spectrometry experiments are highly biased by several factors, such as sample processing, the technician, manufacturer and production of the device (chromatography column, mass-spectrometer), the reagents etc., working in multiple batches introduces further variance in the protein data. This variance would overshadow the biological differences across sample types. Before attempting to remove batch effects it is recommended to elucidate the variance with a dimensionality reduction technique which will be explained in the next chapter. To remove the technical factor also called non-biological experimental variation the R package sva \citep{Leek2012} with the ComBat function \citep{Johnson2007} was used. The expression of a gene $g$ for the batch $m_i$ and sample $j$ is defined
